Conditional Power and Friends: The Why and How of (Un)planned, Unblinded Sample Size Recalculations in Confirmatory Trials

### 2021-08-13 latex/manuscript.tex
...
major refactor

\renewcommand{\Pr}{\operatorname{Pr}}
\newcommand{\E}{\ensuremath{\operatorname{E}}}
\newcommand{\V}{\operatorname{V}}
\newcommand{\st}{\operatorname{subject\ to:\ }}
\newcommand{\argmin}[1]{\operatornamewithlimits{argmin:\ }_{#1}}
\newcommand{\argmax}[1]{\operatornamewithlimits{argmax:\ }_{#1}}

@@

%\begin{comment}
\author{
    Kevin Kunzmann \thanks{
        MRC Biostatistics Unit, University of Cambridge,
        Cambridge Institute of Public Health,
        Forvie Site, Robinson Way,
        Cambridge Biomedical Campus,
        Cambridge CB2 0SR,
        United Kingdom,
        \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk}
    }\\
    MRC Biostatistics Unit \\
    University of Cambridge \\
    Michael J. Grayling \\
    Population Health Sciences Institute \\
    Newcastle University \\
    Kim May Lee \\
    Pragmatic Clinical Trials Unit\\
    Queen Mary University of London\\
    David S.\ Robertson \\
    MRC Biostatisics Unit\\
    University of Cambridge\\
    Kaspar Rufibach \\
    Methods, Collaboration, and Outreach Group (MCO) \\
    Department of Biostatistics\\
    F. Hoffmann-La Roche, Basel\\
    James M. S. Wason \\
    Population Health Sciences Institute \\
    Newcastle University\\
    and \\
     MRC Biostatistics Unit \\
    University of Cambridge
}
%\end{comment}
%\author{}

@@

\begin{abstract}
Adapting the final sample size of a trial to the evidence accruing
during the trial is a natural way to address planning uncertainty.
Designs with adaptive sample size need to
account for their optional stopping to guarantee
strict type-I error-rate control.
A variety of different methods to maintain type-I error-rate control
after unplanned changes of the initial sample size have been proposed
in the literature.
This makes interim analyses for the purpose of sample size 
recalculation feasible in a regulatory context.
Since the sample size is usually determined via an
argument based on the power of the trial,
an interim analysis raises the question of
how the final sample size should be determined conditional on the accrued information.
Conditional power is a concept often put forward in this context.
Since it depends on the unknown effect size,
we take a strict estimation perspective and compare
assumed conditional power, observed conditional power,
and predictive power with respect to their properties as estimators
of the unknown conditional power.
We then demonstrate that pre-planning an interim analysis using
methodology for unplanned interim analyses is ineffective and
naturally leads to the concept of optimal two-stage designs.
We conclude that unplanned design adaptations should only be conducted
as reaction to trial-external new evidence, operational needs to violate the originally chosen design,
or \textit{post~hoc} changes in the objective criterion.
Finally, we show that commonly discussed sample size recalculation rules can lead to paradoxical outcomes and propose two alternative ways of reacting to
newly emerging trial-external evidence.
\end{abstract}

@@

The planning phase of confirmatory clinical trials is typically
characterized by substantial uncertainty about the magnitude of the
parameters underlying the hypothesis of interest.
Often the alternative hypothesis is the superiority of a new intervention arm
over a treatment-as-usual control arm and is formalized via
the average treatment effect, e.g., the difference of means between the
two randomly allocated treatment arms.
Clinical data collection is expensive and time consuming leading to a
strong economic incentive to reach the study goals with as little data as possible.
The conventional statistical criteria to determine the sample size of a trial
are a one-sided type-I error rate of
2.5\% and a power of 80\% or 90\%.
Since power is a function of the unknown effect size the initial design
must be specified under substantial uncertainty about the magnitude of the effect which the trial hopes to detect eventually.
Mainly, two ways of addressing this challenge have been put forward in the literature.

Firstly, the so-called `hybrid' approach to sample size
derivation takes a Bayesian
view on determining the initial sample size of a clinical trial~\cite{spiegelhalter1994}.
The Bayesian component requires the specification of an informative
prior on the parameters of interest.
This then allows reasoning about the `expected power' of a trial as a
function of its sample size and, consequently, determination of the sample size
such that the expected power exceeds a target threshold.
This concept is a straight-forward extension of the usual practice of
computing the power under a fixed point alternative, which can easily be recovered as a special case when considering a point prior.
The advantage lies in the fact that the \textit{a~priori} information about the magnitude of the effect size is faithfully
reflected in the sample size derivation.
Since the actual analysis is still frequentist, type-I error rate control is not compromised.

Secondly, authors have proposed to apply the concept of adaptive design changes
to recalculate the initial sample size of a design based on data observed
within the trial itself~\cite{bauer2016}.
The rationale behind this approach is to use the accruing evidence about the unknown parameters driving the sample size derivation to
`correct' the sample size mid-trial.
During the interim analysis, the accrued data can either be unblinded
or not.
Methods for the latter `blinded interim analysis' are particularly useful if relevant nuisance parameters such as the variance are also unknown \cite{birkett1994internal,bauer2016}.
We focus on the unblinded case which allows for a more precise
interim assessment of the effect size than methods that retain
the blind.
The maximal type-I error rate constraint is then usually protected by
applying the conditional error principle~\cite{muller2004,brannath2012}
or an equivalent formulation via $\alpha$-spending or
combination functions~\cite{bauer2016}.
Often the sample size of the current
trial is adjusted such that the conditional power given the data observed up to the
interim analysis again exceeds the initial threshold for unconditional
power~\cite{proschan1995}.
Any recalculation based on conditional power arguments must
address the problem that conditional power,
just as unconditional power,
depends on the unknown underlying parameters.
It must thus be
estimated to inform a sample size recalculation.
Yet, precise estimation of the relevant parameters before the conclusion of a trial is hard since only a fraction of the final sample size is available.
Three approaches to estimating conditional power have been discussed in the literature.

Firstly,
`conditional power' is the probability to reject the null hypothesis given the interim data as a function of the unknown parameters.
In a slight abuse of terminology, the same term is often used to refer to
the assumed conditional power that is obtained by plugging in the
point alternative used for the initial sample size derivation.
Evidently, this assumed conditional makes no use of the accrued trial
data since the effect size is kept fixed.
Secondly,
authors have proposed to use the `observed conditional power' instead, which replaces the unknown parameters with their maximum-likelihood estimates.
The latter `plug-in' approach is often criticized for
`... [using] the interim estimate of the effect [...] twice ...' Bauer~\textit{et~al.},~p.~330~\cite{bauer2016} and \cite{bauer2006}.
Thirdly,
conditional power can be evaluated as Bayesian expected power conditional on the observed interim data,
i.e., by averaging conditional power as function of the unknown
parameters with respect to the posterior density
after conducting the interim analysis.
Within the hybrid Bayesian framework this is usually referred to as
`predictive power'~\cite{spiegelhalter1994,bauer2016}.

The purpose of this paper is to argue that pre-planned sample size adaptations
using methodology intended for unplanned interim analyses to react to within-trial interim data are inefficient and unnecessary if the original design was planned optimally.
We then discuss cases where an unplanned design adaptation might still be warranted and propose two criteria that allow such an adaptation that is
consistent with the original design.
We first review assumed \mbox{conditional-,} observed \mbox{conditional-,} and predictive power
from an estimation perspective.
For sake of simplicity,
we use a simple single-arm one-stage design to illustrate core
characteristics.
We then discuss the risks of constructing an
`adaptive' two-stage design by na\"{\i}vely applying
a conditional-power-based sample size recalculation rule and the
conditional error principle.
The drawbacks of this na\"{\i}ve approach directly lead to the
concept of optimal-two stage designs and
we build on ideas presented in \cite{brannath2004,pilz2019}, and
\cite{adoptrjss2020} to derive optimal designs for the situation studied in
this manuscript.
Without loss of generality, we focus on minimal expected sample size as optimality criterion.
Optimal two-stage designs, by definition, cannot be made more
efficient by sample size recalculation.
The optimality of such designs does, however, depend on the initial trial-external evidence that feeds into the planning assumptions.
This trial-external evidence might change during an ongoing study
and may thus mandate a sample size recalculation.
In Section~\ref{sec:unplanned-adaptation},
we discuss when an unplanned sample size recalculation is
reasonable,
how such a recalculation interacts with the concept of
optimal two-stage designs,
and the practical implementation.

In the following,
we assume that the interest lies in testing a new treatment for efficacy.
For the sake of simplicity,
we consider the case of a single-arm trial.
All considerations can easily be extended to the more
practically relevant two-arm case.
We further assume that the individually observed outcomes of the study participants $X_i, i = 1,\ldots,n$ are $iid$ and that their distribution has finite first moment $\theta$ and unit variance.
Again, all considerations can be extended to the case of
generic known variances and,
at least approximately, to the case of unknown variance.
A suitable test statistic for the null hypothesis of interest~${\mathcal{H}_0:\theta\leq0}$ is
\begin{align}
    Z_n = \frac{1}{\sqrt{n}} \sum_{i=1}^n X_i \ .
\end{align}
Invoking the central limit theorem,
$Z_n\stackrel{\cdot}{\sim}\mathcal{N}(\sqrt{n}\,\theta, 1)$
and,
on the boundary of the null hypothesis, $Z_n\stackrel{\cdot}{\sim}\mathcal{N}(0, 1)$.

Further assuming a maximal permissible type-I error rate of $\alpha=0.025$ (which we use for the remainder of the paper), the critical value for a single-stage fixed-sample-size design is the
$1-\alpha$ quantile of the standard normal distribution, i.e., approximately $c=1.96$.
The test then rejects $\mathcal{H}_0$ if and only if $Z_n > c$ after the outcomes of $n$ subjects have been observed.
The required size of the trial $n$, given a maximal allowable type-I error
rate, is usually determined by some form of restriction on
the (minimal) statistical power of the test.
Approaches to defining such a power constraint under
different assumptions were reviewed in~\cite{kunzmann2020}.

\section{Monitoring power} %%%%%%%%%%%%%%%%%%%%%%%
\label{sec:monitoring-power}

No matter which rationale was used to derive the overall sample size $n$,
after observing $m$, $0<m<n$, outcomes,
an independent data monitoring committee might be interested to learn about the prospects of
eventually rejecting the null hypothesis.
Both the rejection probability conditional on the presence of an effect (conditional power) or the joint probability of rejection and presence of an effect (conditional probability of success) could be of interest.
Here, we focus on the former due to the central role of
power (and conditional power) in the planning of clinical trials.
This effectively constitutes an estimation problem where the estimand is
the \emph{conditional} probability of rejecting the null hypothesis given
$Z_m=z_m$ and $\theta > 0$.
Here the conditioning on a positive effect is crucial to justify the notion
of `power' in contrast to the unconditional `probability of success'~\cite{kunzmann2020}.

Since $Z_n$ and $Z_m$ are (asymptotically) jointly normal, the conditional distribution of $Z_n$ given $Z_m$ and $\theta$ is again normal and
given by
\begin{align}
    \mathcal{L}_\theta\big[\, Z_n \mid Z_m = z_m \,\big] \stackrel{\cdot}{=} \mathcal{N}\Big(\sqrt{n}\,\theta + \sqrt{\tau}\big(z_{m} - \sqrt{m}\,\theta\big),\ 1 - \tau\Big) \label{eq:joint-distribution}
\end{align}
where $\tau:=m/n$ is the `information fraction' at the interim analysis.

The probability of rejecting the null hypothesis at the end of the trial
as a function of $\theta$ is referred to as the `conditional power' in the literature.
It is defined as
\begin{align}
    \CP(z_m, c, \theta) :&= \Pr_\theta[\,Z_n > c \cond Z_m = z_m\,] \\
        &= 1 - \Phi\left(
            \frac{c - \sqrt{n}\,\theta - \sqrt{\tau} \, z_m + \sqrt{\tau}\sqrt{m}\,\theta
            }{\sqrt{1 - \tau}}
        \right) \\
        &= 1 - \Phi\left(
            \frac{c/\sqrt{m} + \sqrt{\tau}\,\big(\theta - \widehat{\theta}_m\big) - \theta/\sqrt{\tau}
            }{\sqrt{(n - m)/(n\,m)}}
        \right)      \ .
\end{align}
Ignoring the implicit dependency on the time point of the interim analysis, $m$,
conditional power is thus a function of the observed effect $\widehat{\theta}_m:=z_m/\sqrt{m}$
(or equivalently the observed test statistic $z_m$), the critical value for the test decision~$c$,
and the standardized effect size~$\theta$.
Since $\theta$ is unknown,
so is $\CP(z_m, c, \theta)$ and it thus cannot be
evaluated directly upon observing $Z_m=z_m$.
Yet, as a function of the unknown quantity $\theta$,
$\CP(z_m, c, \theta)$ can be estimated from observed data.
The estimation perspective on `evaluating' conditional power is
less commonly taken in the literature but provides a consistent framework to
compare the characteristics of different methods~\cite{bauer2006}.

Firstly, conditional power can be estimated based on a fixed point alternative~$\theta_1$.
In a slight abuse of terminology,
this quantity is often also referred to as `conditional power'.
To clearly distinguish it from $\CP(z_m, c, \theta)$ we denote
it `assumed conditional power'
\begin{align}
    \ACP(z_m,c) := \CP\big(\, z_m, c, \theta_1 \,\big) \ .
\end{align}

Secondly, the observed effect $\widehat{\theta}_m$ can be used
as a plug-in estimator for the unknown effect size~\cite{proschan1995}.
This quantity is sometimes referred to as `observed conditional power'
in the literature and is defined as
\begin{align}
    \OCP(z_m, c) := \CP\big(\, z_m, c, \widehat{\theta}_m \,\big) \ .
\end{align}

Thirdly, a Bayesian approach can be taken if one
is willing to quantify the uncertainty about the unknown parameter $\theta$ by modeling it as the
realization of a random variable $\Theta\sim\varphi(\cdot)$ where $\varphi(\theta)$ is the
prior probability density function evaluated at the parameter value $\theta$.
Our definition of this so-called `predictive power'
\begin{align}
    \PP_\varphi(z_m,c) :&= \Pr_\varphi[\,Z_n > c \cond \Theta >0, Z_m = z_m\,] \\
    &= \int_{0}^{\infty}\CP\big(z_m, c, \theta\big)\,\varphi\big(\theta\cond Z_m=z_m, \Theta\geq0\big)\operatorname{d}\theta
\end{align}
differs slightly from the one proposed by~\cite{spiegelhalter1994}
in that we condition on a positive effect size $\Theta>0$.
This is more consistent with the notion of frequentist power as discussed in~\cite{kunzmann2020}, although the difference is
only of practical relevance when a substantial
fraction of the \textit{a~priori} probability mass is concentrated on the null hypothesis.

We now compare $\ACP, \OCP$, and $\PP$ by means of a concrete example.
To this end we assume that the available prior information can be summarized in a truncated normal prior with density
\begin{align}
    \varphi(\theta) := \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\frac{\displaystyle \phi\left(\frac{\theta - 0.4}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - 0.4}{0.2}\right) - \Phi\left(\frac{-0.5 - 0.4}{0.2}\right)} \ ,
\end{align}
(see Figure~\ref{fig:prior-pdf})
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/prior_density}
    \caption{%
        Assumed prior density function.
        The gray area indicates the null hypothesis, $\mathcal{H}_0$, of no effect.
    }
    \label{fig:prior-pdf}
\end{figure}
and that all positive effect sizes are clinically relevant.
A truncated normal prior is convenient since it allows the analytic
computation of the posterior distribution under a normal likelihood.
It is also the maximum entropy distribution on a compact interval given mean and standard deviation and thus `least informative'
given a lower and upper boundaries on plausible effects as well as
the location and vagueness (variance) of the prior.
Following~\cite{kunzmann2020} the required
sample size $n$ is then determined by requiring
a minimal expected power of $1-\beta=80\%$,
where expected power is defined as
\begin{align}
    \EP_\varphi(n, c) :=
    \Pr\big[\,Z_n>c\,|\,\Theta\geq0\,\big] = \int \Pr_\theta\big[\,Z_n>c\,\big]\, \varphi(\theta\,|\,\Theta\geq0) \,\operatorname{d}\theta \ .
\end{align}
Under these assumptions, the required sample size is
$n = 79$.
No point alternative is necessary to derive this sample size.
We thus use the prior mean rounded to the first decimal point to
evaluate $\ACP$ and define $\theta_1:=0.4$.

>>>>

\renewcommand{\Pr}{\operatorname{Pr}}
\newcommand{\E}{\ensuremath{\operatorname{E}}}
\newcommand{\V}{\operatorname{V}}
\newcommand{\solve}[1]{\operatornamewithlimits{solve:\ }_{#1}}
\newcommand{\st}{\operatorname{subject\ to:\ }}
\newcommand{\argmin}[1]{\operatornamewithlimits{argmin:\ }_{#1}}
\newcommand{\argmax}[1]{\operatornamewithlimits{argmax:\ }_{#1}}
\newcommand{\etal}{\textit{et~al.}}

@@

\author{
    \footnotesize Kevin Kunzmann\\
    \footnotesize MRC Biostatistics Unit, University of Cambridge \\
    \footnotesize
    East Forvie Building, Robinson Way,
    Cambridge Biomedical Campus \\
    \footnotesize Cambridge CB2 0SR,
    United Kingdom \\
    \footnotesize \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk}
    \and
    \footnotesize Michael J. Grayling \\
    \footnotesize Population Health Sciences Institute, Newcastle University
    \and
    \footnotesize Kim May Lee \\
    \footnotesize Institute of Psychiatry, Psychology and Neuroscience, King's College London
    \and
    \footnotesize David S.\ Robertson \\
    \footnotesize MRC Biostatisics Unit, University of Cambridge
    \and
    \footnotesize Kaspar Rufibach \\
    \footnotesize Methods, Collaboration, and Outreach Group (MCO) \\
    \footnotesize Department of Biostatistics, F. Hoffmann-La Roche, Basel
    \and
    \footnotesize James M. S. Wason \\
    \footnotesize Population Health Sciences Institute, Newcastle University and \\
    \footnotesize MRC Biostatistics Unit, University of Cambridge
}

@@

\begin{abstract}
Adapting the final sample size of a trial to the evidence accruing
during the trial is a natural way to address planning uncertainty.
Since the sample size is usually determined by an
argument based on the power of the trial,
an interim analysis raises the question of
how the final sample size should be determined conditional on the accrued information.
To this end, we first review and compare common approaches to estimating conditional power
which is often used in heuristic sample size recalculation rules.
We then discuss the connection of heuristic sample size recalculation and
optimal two-stage designs demonstrating that the latter is the superior approach in a fully pre-planned setting.
Hence, unplanned design adaptations should only be conducted
as reaction to trial-external new evidence, operational needs to violate the originally chosen design,
or \textit{post~hoc} changes in the optimality criterion but not as a reaction to trial-internal data.
We are able to show that commonly discussed sample size recalculation rules lead to paradoxical adaptations
where an initially planned optimal design is not invariant under the adaptation rule
even if the planning assumptions do not change.
Finally, we propose two alternative ways of reacting to
newly emerging trial-external evidence in ways that are consistent with the originally
planned design to avoid such inconsistencies.
\end{abstract}

@@

The planning phase of confirmatory clinical trials is typically
characterized by substantial uncertainty about the magnitude of the
parameters underlying the hypothesis of interest.
Clinical data collection is expensive and time consuming leading to a
strong economic incentive to reach the study goals with as little data as possible.
The conventional statistical criteria to determine the sample size of a trial
are a one-sided type~I error rate of
2.5\% and a power of 80\% or 90\%.
Since power is a function of the unknown effect size, the initial design
must be specified under substantial uncertainty.
Mainly, two ways of addressing this challenge have been put forward in the literature.

Firstly, the so-called "hybrid" approach to sample size
derivation takes a Bayesian
view on determining the initial sample size of a clinical trial~\cite{spiegelhalter1994},
which requires the specification of an informative
prior on the parameters of interest.
This then allows reasoning about the "expected power" of a trial as a
function of its sample size and, consequently, determination of the sample size
such that the expected power exceeds a target threshold.
This concept is a straight-forward extension of the usual practice of
computing the power under a fixed point alternative, which can be recovered as a special case when considering a point prior.
The advantage lies in the fact that the \textit{a~priori} information about the magnitude of the effect size is faithfully
reflected in the sample size derivation.
Since the actual analysis is still frequentist, type~I error rate control is not compromised.

Secondly, authors have proposed to apply the concept of adaptive design changes
to recalculate the initial sample size of a design based on data observed
within the trial itself~\cite{bauer2016}.
The rationale behind this approach is to use the accruing evidence about the unknown parameters driving the sample size derivation to
"correct" the sample size mid-trial.
During the interim analysis, the accrued data can either be unblinded
or not.
Methods for the latter blinded interim analysis are particularly useful if relevant nuisance parameters such as the variance are also unknown \cite{birkett1994internal,bauer2016}.
We focus on the unblinded case which allows for a more precise
interim assessment of the effect size than methods that retain
the blind.
The maximal type~I error rate constraint is then usually protected by
applying the conditional error principle~\cite{muller2004,brannath2012}
or an equivalent formulation via $\alpha$-spending or
combination functions~\cite{bauer2016}.
Often the sample size of the current
trial is adjusted such that the conditional power given the data observed up to the
interim analysis again exceeds the initial threshold for unconditional
power~\cite{proschan1995}.
Any recalculation based on conditional power arguments must
address the problem that conditional power,
just as unconditional power,
depends on the unknown underlying parameters
and must thus be
estimated to inform a sample size recalculation.
Yet, precise estimation of the relevant parameters before the conclusion of a trial is hard since only a fraction of the final sample size is available.
Three approaches to estimating conditional power have been discussed in the literature.

Conditional power is the probability to reject the null hypothesis given the interim data as a function of the unknown parameters.
Firstly, in a slight abuse of terminology, the same term is often used to refer to
the assumed conditional power that is obtained by plugging in the
point alternative used for the initial sample size derivation.
Evidently, this assumed conditional makes no use of the accrued trial
data since the effect size is kept fixed.
Secondly,
authors have proposed to use the "observed conditional power" instead, which replaces the unknown parameters with their maximum-likelihood estimates.
The latter approach is often criticized for
`... [using] the interim estimate of the effect [...] twice ...' Bauer~\textit{et~al.},~p.~330~\cite{bauer2016} and Bauer and K\"onig\cite{bauer2006}.
Thirdly,
conditional power can be evaluated as Bayesian expected power conditional on the observed interim data,
i.e., by averaging conditional power as function of the unknown
parameters with respect to the posterior density
after conducting the interim analysis.
Within the hybrid Bayesian framework this is usually referred to as
"predictive power"~\cite{spiegelhalter1994,bauer2016}.

The purpose of this paper is to argue that pre-planned sample size adaptations
using methodology intended for unplanned interim analyses to react to within-trial interim data are inefficient and unnecessary if the original design was planned optimally and should thus be avoided.
We discuss cases where an unplanned design adaptation might still be warranted as reaction to emerging trial-external evidence and
propose two novel adaptation methods that are consistent with the original design in the sense that the original design is only changed when the planning assumptions are modified.

To this end, we first review assumed \mbox{conditional-,} observed \mbox{conditional-,} and predictive power
from an estimation perspective.
We then discuss the risks of constructing an
adaptive two-stage design by na\"{\i}vely applying
a conditional-power-based sample size recalculation rule and the
conditional error principle.
The drawbacks of this na\"{\i}ve approach directly lead to the
concept of optimal-two stage designs \cite{brannath2004,pilz2019,adoptrjss2020}.
Without loss of generality, we focus on minimal expected sample size as optimality criterion.
These optimal two-stage designs, by definition, cannot be made more
efficient by sample size recalculation.
The optimality of such designs does, however, depend on the initial trial-external evidence that feeds into the planning assumptions.
This trial-external evidence might change during an ongoing study
and may thus still mandate a sample size recalculation.

In Section~\ref{sec:unplanned-adaptation},
we then discuss,
how such a recalculation interacts with the concept of
optimal two-stage designs.
In particular, we demonstrate that commonly used methods
are inconsistent in that they imply that that the original design needs to be modified for every possible interim outcome even if there is no change to the planning assumptions.
We propose two novel methods that overcome this inconsistency.

In the following,
we assume that the interest lies in testing a new treatment for efficacy.
For the sake of simplicity,
we consider the case of a single-arm trial.
All considerations can easily be extended to the two-arm case (see the application example in Section~\ref{2arm}).
We further assume that the individually observed outcomes of the study participants $X_i, i = 1,\ldots,n$ are $iid$ and that their distribution has finite first moment $\theta$ and unit variance.
Again, all considerations can be extended to the case of
generic known variances and,
at least approximately, to the case of unknown variance.
A suitable test statistic for the null hypothesis of interest~${\mathcal{H}_0:\theta\leq0}$ is
\begin{align}
    Z_n = \frac{1}{\sqrt{n}} \sum_{i=1}^n X_i \ .
\end{align}
Invoking the central limit theorem,
$Z_n\stackrel{\cdot}{\sim}\mathcal{N}(\sqrt{n}\,\theta, 1)$
and,
on the boundary of the null hypothesis, $Z_n\stackrel{\cdot}{\sim}\mathcal{N}(0, 1)$.
Further assuming a maximal permissible type~I error rate of $\alpha=0.025$ (which we use for the remainder of the paper), the critical value for a single-stage fixed-sample-size design is the
$1-\alpha$ quantile of the standard normal distribution, i.e., approximately $c=1.96$.
The test then rejects $\mathcal{H}_0$ if and only if $Z_n > c$ after the outcomes of $n$ subjects have been observed.
The required size of the trial $n$, given a maximal allowable type~I error
rate, is usually determined by some form of restriction on
the (minimal) statistical power of the test.
Approaches to defining such a power constraint under
different assumptions were reviewed in~Kunzmann~\etal\cite{kunzmann2020}.

\section{Previous work}

\subsection{Monitoring power} %%%%%%%%%%%%%%%%%%%%%%%
\label{sec:monitoring-power}

No matter which rationale was used to derive the overall sample size $n$,
after observing $m$, $0<m<n$, outcomes,
an independent data monitoring committee might be interested to learn about the prospects of
eventually rejecting the null hypothesis.
Since $Z_n$ and $Z_m$ are (asymptotically) jointly normal, the conditional distribution of $Z_n$ given $Z_m$ and $\theta$ is again normal and
given by
\begin{align}
    \mathcal{L}_\theta\big[\, Z_n \mid Z_m = z_m \,\big] \stackrel{\cdot}{=} \mathcal{N}\Big(\sqrt{n}\,\theta + \sqrt{\tau}\big(z_{m} - \sqrt{m}\,\theta\big),\ 1 - \tau\Big) \label{eq:joint-distribution}
\end{align}
where $\tau:=m/n$ is the "information fraction" at the interim analysis (and the squared correlation between $Z_n$ and $Z_m$).
The probability of rejecting the null hypothesis at the end of the trial
as a function of $\theta$ is referred to as the conditional power in the literature.
In the setting at hand, it is defined as
\begin{align}
    \CP(z_m, c, \theta) :&= \Pr_\theta[\,Z_n > c \cond Z_m = z_m\,] \\
        &= 1 - \Phi\left(
            \frac{c - \sqrt{n}\,\theta - \sqrt{\tau} \, z_m + \sqrt{\tau}\sqrt{m}\,\theta
            }{\sqrt{1 - \tau}}
        \right) \\
        &= 1 - \Phi\left(
            \frac{c/\sqrt{m} + \sqrt{\tau}\,\big(\theta - \widehat{\theta}_m\big) - \theta/\sqrt{\tau}
            }{\sqrt{(n - m)/(n\,m)}}
        \right)      \ .
\end{align}
Since $\theta$ is unknown,
so is $\CP(z_m, c, \theta)$ and it cannot be
evaluated directly upon observing $Z_m=z_m$.
Yet, as a function of the unknown quantity $\theta$,
$\CP(z_m, c, \theta)$ can be estimated from observed data.
The estimation perspective on "evaluating" conditional power is
less commonly taken in the literature but provides a consistent framework to
compare the characteristics of different methods~\cite{bauer2006}.

Firstly, conditional power can be estimated based on a fixed point alternative~$\theta_1$.
In a slight abuse of terminology,
this quantity is often also referred to as "conditional power".
To clearly distinguish it from $\CP(z_m, c, \theta)$ we denote
it "assumed conditional power"
\begin{align}
    \ACP(z_m,c) := \CP\big(\, z_m, c, \theta_1 \,\big) \ .
\end{align}

Secondly, the observed effect $\widehat{\theta}_m$ can be used
as a plug-in estimator for the unknown effect size~\cite{proschan1995}.
This quantity is sometimes referred to as "observed conditional power"
in the literature and is defined as
\begin{align}
    \OCP(z_m, c) := \CP\big(\, z_m, c, \widehat{\theta}_m \,\big) \ .
\end{align}

Thirdly, a Bayesian approach can be taken if one
is willing to quantify the uncertainty about the unknown parameter $\theta$ by modeling it as the
realization of a random variable $\Theta\sim\varphi(\cdot)$ where $\varphi(\theta)$ is the
prior probability density function evaluated at the parameter value $\theta$.
Our definition of this so-called "predictive power"
\begin{align}
    \PP_\varphi(z_m,c) :&= \Pr_\varphi[\,Z_n > c \cond \Theta >0, Z_m = z_m\,] \\
    &= \int_{0}^{\infty}\CP\big(z_m, c, \theta\big)\,\varphi\big(\theta\cond Z_m=z_m, \Theta>0\big)\operatorname{d}\theta
\end{align}
differs slightly from the one proposed by~Spiegelhalter~\etal\cite{spiegelhalter1994}
in that we condition on a positive effect size $\Theta>0$.
This is more consistent with the notion of expected power as discussed in~Kunzmann~\etal\cite{kunzmann2020}
\begin{align}
    \EP_\varphi(n, c) :&=
    \Pr_\varphi\big[\,Z_n>c\,|\,\Theta>0\,\big]\\
    &= \int \PP_\varphi(z_m,c)\ f_\varphi(z_m\mid \Theta>0) \,\operatorname{d}z_m
\end{align}
where $_\varphi(z_m\mid \Theta>0)$ is the probability density function of the predictive distribution of $Z_m\mid \Theta>0$.
The difference is
only of practical relevance when a substantial
fraction of the \textit{a~priori} probability mass is concentrated on the null hypothesis.

We now compare $\ACP, \OCP$, and $\PP$ by means of a concrete example.
To this end we assume that the available prior information can be summarized in a truncated normal prior with density
\begin{align}
    \varphi(\theta) := \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\frac{\displaystyle \phi\left(\frac{\theta - 0.4}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - 0.4}{0.2}\right) - \Phi\left(\frac{-0.5 - 0.4}{0.2}\right)} \ ,
\end{align}
(see Figure~\ref{fig:prior-pdf})
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/prior_density}
    \caption{%
        Assumed prior density function.
        The gray area indicates the null hypothesis, $\mathcal{H}_0$, of no effect.
    }
    \label{fig:prior-pdf}
\end{figure}
and that all positive effect sizes are clinically relevant.
A truncated normal prior is convenient since it allows the analytic
computation of the posterior distribution under a normal likelihood.
It is also the maximum entropy distribution on a compact interval given mean and standard deviation and thus "least informative"
given a lower and upper boundary on plausible effects as well as
the location and vagueness (variance) of the prior.
The choice of the prior is a key consideration, but an extensive discussion is beyond the scope of this paper. We refer the reader to discussion in Spiegelhalter \textit{et al.}~\cite{spiegelhalter2004,rufibach_16}, and Hampson \textit{et al.}~\cite{hampson2021}, as well as the formal prior elicitation framework SHELF~\cite{kinnersley2013, shelf}, which is routinely used in practice by pharmaceutical companies~\cite{dallow2018}.

Following~\cite{kunzmann2020} the required
sample size $n$ is determined by requiring
a minimal expected power of $1-\beta=80\%$.
Under these assumptions, the required sample size is
$n = 79$.
Since no point alternative is necessary to derive this sample size, we use the prior mean rounded to the first decimal point to
evaluate $\ACP$ and define $\theta_1:=0.4$.

The three estimators are depicted in Figure~\ref{fig:acp-ocp-pp-comparison-absolute} as functions
of the observed interim outcome, where the interim time-point $m = 26$ (chosen heuristically as approximately 1/3 of the overall sample size).
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_example}
        \caption{%
            Estimates as function of the interim data.
        }
        \label{fig:acp-ocp-pp-comparison-absolute}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_sampling_distributions}
        \caption{%
            Histograms of the sampling distributions.
        }
        \label{fig:sampling-distributions}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_bias_mae_mse}
        \caption{%
            Bias, mean absolute error (MAE) and mean squared error (MSE).
        }
        \label{fig:acp-ocp-pp-as-estimators}
    \end{subfigure}
    \caption{%
        Properties of $\ACP,\OCP$, and $\PP$ as estimators of the unknown conditional power at $m=26$ and overall sample size $n=79$.
    }
\end{figure}
The most sensitive measure is $\OCP$ while $\ACP$ is the least sensitive
(see also spread of the sampling distributions in Figure~\ref{fig:sampling-distributions}).
This is due to $\ACP$ assuming a fixed effect size and thus
only being affected by direct changes in $z_m$.
Both $\OCP$ and $\PP$, however, are also indirectly affected by updating
the belief about the effect size with the interim results.
The plug-in estimate $\OCP$ assumes that the observed effect is the
true effect and does not quantify the uncertainty around this value in any way.
$\PP$, on the other hand, invokes Bayes' theorem and then integrates
$\CP(z_m,c,\theta)$ with respect to the obtained posterior distribution.
The degree of adaptation of $\PP$ thus depends
on the vagueness of the chosen prior.
In fact, as the prior approaches a point mass at $\theta_1$, $\PP$ converges
to $\ACP$.
$\ACP$ can thus be seen as a mere special case of $\PP$ with a point prior
on $\theta_1>0$.

A fundamental difference between $\ACP$ and $\PP$ on the one hand and $\OCP$ on the other hand is that
$\OCP$ is the only estimator that does not condition on~$\Theta$.
In contrast, $\ACP$ implicitly conditions on $\Theta=\theta_1>0$ and
$\PP$~on~${\Theta > 0}$.
Since $\OCP$ does not condition on $\Theta>0$, the sampling distribution is much more left-skewed for small effects than for the other two estimators (see~Figure~\ref{fig:sampling-distributions}).
Also, the negligence of the sampling variation of $\widehat{\theta}_m$, which $\OCP$ plugs in the expression for conditional power, leads to a larger variance for intermediate values of $\theta$ and the characteristic
U-shape \cite{bauer2006}.
This high variance of $\OCP$ directly translates to a relatively large mean absolute error (MAE) and mean squared error (MSE) when estimating conditional power,
see Figure~\ref{fig:acp-ocp-pp-as-estimators}.
$\PP$ is the posterior expectation of conditional power with respect to the chosen prior (conditional on a positive effect) and thus  minimizes the quadratic Bayes risk, i.e.~the average MSE weighted by $\varphi(\cdot\mid\Theta>0)$.
This leads to relative good precision for parameter values with high \textit{a~priori} likelihood
(around $\theta=0.4$).
If one wanted to minimize the MAE directly,
the posterior median would be optimal.
The principle, however, remains the same and
the posterior mean is more consistent with the derivation of the initial sample size using expected power.
$\ACP$ is clearly the best in terms of bias and MAE/MSE for values close to or
above $\theta_1$, but its performance as estimator of the conditional power
quickly deteriorates for small effect sizes.
